Verifying Concurrent Search Structure Templates for C Programs using VST

### 2023-04-05 concurrency/CONCUR 2023/paper.tex
WIP

\section{Introduction}
\label{sec:introduction}
Krishna et al. proposed concurrent search structure templates~\cite{templates} as a method for separating the proof of correctness of a concurrent access method (optimistic concurrency, hand-over-hand locking, and internal links) from the proof of correctness of the underlying data structure (linked list, hashtable, and B-tree). Ideally, it should be possible to prove the correctness of $n$ (single-threaded) data structure implementations and $m$ concurrency patterns and obtain $n \times m$ verified concurrent data structures. However, in practice, the story is more complicated: certain patterns work only for specific data structures or require the data structures to store extra information, while some internal data structure operations may not fit the template model. In this paper, we apply the template approach to concurrent data structure implementations in C, which we verify using the Verified Software Toolchain (VST)~\cite{plfcc}, and report on its effectiveness and the challenges we encountered. The template approach depends crucially on the idea of \emph{logically atomic specifications} introduced in TaDA~\cite{tada} and further developed in Iris~\cite{iris}; our proofs take advantage of recent work integrating Iris-style logical atomicity into VST~\cite{mansky2022bringing}.

\section{Background}
\subsection{Concurrent Search Structure Templates}
A search structure is a type of data structure that is designed to facilitate the efficient storage and retrieval of data based on specific search criteria. These structures organize data in a manner that enables quick and efficient search, insertion, deletion, and traversal operations. Concurrent search structures, in turn, are data structures that can be accessed and modified concurrently by multiple threads or processes. The design of concurrent search structures poses several significant challenges, including the need to ensure correctness and consistency under concurrent access, achieve scalability by minimizing contention and maximizing parallelism, maintain performance and efficiency while managing synchronization and memory, and ensure compatibility with diverse hardware and software platforms. The complex interplay between concurrency and shared memory in concurrent search structures makes formal verification a challenging task. Modularity is a crucial aspect of simplifying formal proofs for concurrent search structures, as it is for designing and maintaining large systems. By decomposing complex systems into smaller, more manageable components, modularity facilitates the separation of concerns and minimizes interactions between different parts of the system, thus reducing the complexity of formal proofs. Moreover, modularity permits the reuse of verified components, which further simplifies the verification process.
{\color{red}{(Say something to link to template algorithms)}}

@@

\section{Search Structure Templates}
The C implementation of a node in our template style is:
\begin{lstlisting}[language = C, backgroundcolor=\color{white},   
	commentstyle=\color{codegreen},
	basicstyle=\ttfamily\footnotesize,
	breakatwhitespace=false,  breaklines=true,  keepspaces=true,                                
	showspaces=false,  showstringspaces=false,
	showtabs=false, tabsize=1]
typedef struct tree {int key; void *value; struct tree_t *left, *right;} tree;
typedef struct tree_t {tree *t; lock_t *lock;} tree_t;
typedef struct pn {struct tree_t *p; struct tree_t *n;} pn;
\end{lstlisting}
Each node (of type \lstinline{tree_t}) contains a \lstinline{lock} field that protects the node and a \lstinline{t} field. The \lstinline{t} field can be either \lstinline{NULL}, indicating that the node is a leaf node with no key or value, or it can point to a \lstinline{tree} struct that stores the node's key, value, and child pointers. 
To implement hand-over-hand locking, which guarantees that at least one lock is maintained during traversal between nodes (i.e., the lock for the next node is obtained before releasing the lock for the current node), the \lstinline{pn} structure must contain two nodes of type \lstinline{tree_t}: the \lstinline{p} field, representing the current node, and the \lstinline{n} field, representing the next node.

\subsection{Lock Coupling}
The lock-coupling template employs the hand-over-hand locking scheme to prevent interference from other threads during traversal. Each thread continuously maintains at least one lock during traversal between nodes, ensuring that other threads cannot interfere with the traversal or invalidate the ongoing search. Figure \ref{traverse_lock} presents two versions of the \lstinline{traverse} function: one written in C (Figure \ref{traverse_lock_a}) and the other in an ML-like language (Figure \ref{traverse_lock_b}). The lock-coupling template algorithm, described in an ML-like language, is presented in Figure \ref{traverse_lock_b} \cite{shasha_1988}.  The algorithm operates on the assumption that certain helper function(s) must be provided by the implementation, which satisfy certain criteria.  Such a function, namely \lstinline{findNext}, is utilized by the algorithm. The \lstinline{findNext} function determines the next node $n'$ to be visited based on the current node $n$ and the key $k$ belonging to the set of key values.

\begin{figure}[!ht]
	\begin{subfigure}[t]{0.48\textwidth}
		\lstinputlisting[language=C, style=myStyle]{lock_traverse.c} 
		\caption{The \lstinline{traverse} method of the lock-coupling template give-up template algorithm written in C}
		\label{traverse_lock_a}
	\end{subfigure}\qquad
	\begin{subfigure}[t]{0.45\textwidth}
		\lstinputlisting[language=caml, style=myStyle]{lock_traverse.ml} 
		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in ML-like language}
		\label{traverse_lock_b}	
	\end{subfigure}
	\caption{The \lstinline{traverse} method of the lock-coupling template algorithms}
	\label{traverse_lock}
\end{figure}

\subsection{Give-Up}

The give-up template does not maintain locks during traversal between nodes; instead, it acquires them only when loading or storing to a node. In each node, the range field serves as a reference point for threads to determine if they are still on the correct path. If a thread finds that it has deviated from the expected path, it abandons the node by releasing the lock and returns to the root node to start again. This behavior aligns with the give-up template algorithm's approach, wherein threads terminate their search upon encountering obstacles or deviating from the expected path.

@@

Figure \ref{traverse_giveup_b} depicts the give-up template algorithm, which is presented in an ML-like language and described in \cite{shasha_1988}. 
Similar to the lock-coupling template algorithm, the give-up template algorithm utilizes a helper function named \lstinline{inRange} in addition to the \lstinline{findNext n k} function discussed earlier in the lock-coupling section. The role of the \lstinline{inRange} function is to check whether the key value $k$ belongs to the range of node $n$. If the key value is outside the node's range, the search is abandoned, and the thread is restarted.

>>>>

\section{Introduction}
\label{sec:introduction}
Krishna et al. proposed concurrent search structure templates~\cite{templates} as a method for separating the proof of correctness of a concurrent access method (optimistic concurrency, hand-over-hand locking, and internal links) from the proof of correctness of the underlying data structure (linked list, hashtable, and B-tree). Ideally, it should be possible to prove the correctness of $n$ (single-threaded) data structure implementations and $m$ concurrency patterns and obtain $n \times m$ verified concurrent data structures. However, in practice, the story is more complicated: certain patterns work only for specific data structures or require the data structures to store extra information, while some internal data structure operations may not fit the template model.

In this paper, we apply the template approach to concurrent data structure implementations in C. We verify them using the Verified Software Toolchain (VST)~\cite{plfcc}, and report on its effectiveness and the challenges we encountered. The template approach depends crucially on the idea of \emph{logically atomic specifications} introduced in TaDA~\cite{tada} and further developed in Iris~\cite{iris}. Our proofs make use of recent work that has integrated Iris-style logical atomicity into VST~\cite{mansky2022bringing}.

\section{Background}
\subsection{Concurrent Search Structure Templates}
A search structure is a type of data structure that is designed to efficiently store and retrieve data based on specific search criteria. These structures organize data in a manner that enables quick and efficient search, insertion, deletion, and traversal operations. Concurrent search structures, in turn, are data structures that can be accessed and modified concurrently by multiple threads or processes. However, the design of concurrent search structures poses several significant challenges, such as ensuring correctness and consistency under concurrent access, achieving scalability by minimizing contention and maximizing parallelism, and maintaining performance and efficiency while managing synchronization and memory. Additionally, these structures must ensure compatibility with diverse hardware and software platforms.

The complex interplay between concurrency and shared memory in concurrent search structures makes formal verification a challenging task. Modularity is a crucial aspect of simplifying formal proofs for concurrent search structures, as it is for designing and maintaining large systems. By decomposing complex systems into smaller, more manageable components, modularity facilitates the separation of concerns and minimizes interactions between different parts of the system, thus reducing the complexity of formal proofs. Moreover, modularity permits the reuse of verified components, which further simplifies the verification process. Concurrent search structures such as lock-free hash tables, concurrent tries, and B-trees are frequently employed. These structures allow developers to create systems that are both efficient and scalable, while also ensuring thread safety.

{\color{red}{(Say something to connect to template algorithms)}}

@@

\section{Search Structure Templates}
The C implementation of a node in our template style is:
\begin{lstlisting}[language = C, backgroundcolor=\color{white},   
	commentstyle=\color{codegreen},
	basicstyle=\ttfamily\footnotesize,
	breakatwhitespace=false,  breaklines=true,  keepspaces=true,                                
	showspaces=false,  showstringspaces=false,
	showtabs=false, tabsize=1]
typedef struct tree {int key; void *value; struct tree_t *left, *right;} tree;
typedef struct tree_t {tree *t; lock_t *lock;} tree_t;
typedef struct pn {struct tree_t *p; struct tree_t *n;} pn;
\end{lstlisting}
Each node of type \lstinline{tree_t} has a \lstinline{lock} field that protects the node and a \lstinline{t} field that points to a \lstinline{tree} struct. The \lstinline{t} field can be either \lstinline{NULL}, indicating that the node is a leaf node with no key or value, or it can point to a \lstinline{tree} struct that stores the node's key, value, and child pointers. The \lstinline{pn} struct contains two nodes of type \lstinline{tree_t}: the \lstinline{p}  field, which denotes the current node, and the \lstinline{n} field, which represents the next node. To traverse the tree using hand-over-hand locking, we require at least one lock is maintained during traversal between nodes, meaning that the lock for the next node is obtained before releasing the lock for the current node.

\subsection{Lock Coupling}
The lock-coupling template employs the hand-over-hand locking scheme to prevent interference from other threads during traversal. Each thread continuously maintains at least one lock during traversal between nodes, ensuring that other threads cannot interfere with the traversal or invalidate the ongoing search. Figure \ref{traverse_lock} presents two versions of the \lstinline{traverse} function: one written in C (Figure \ref{traverse_lock_a}) and the other in an ML-like language (Figure \ref{traverse_lock_b}). The lock-coupling template algorithm, described in an ML-like language, is presented in Figure \ref{traverse_lock_b} \cite{shasha_1988}.  The algorithm assumes that certain helper function(s) must be provided by the implementation that satisfy certain criteria.  One such function, namely \lstinline{findNext}, is utilized by the algorithm to determine the next node $n'$ to be visited based on the current node $n$ and the key $k$ belonging to the set of key values.

\begin{figure}[ht]
	\begin{subfigure}[t]{0.48\textwidth}
		\lstinputlisting[language=C, style=myStyle]{lock_traverse.c} 
		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in C}
		\label{traverse_lock_a}
	\end{subfigure}\qquad
	\begin{subfigure}[t]{0.45\textwidth}
		\lstinputlisting[language=caml, style=myStyle]{lock_traverse.ml} 
		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in ML-like language}
		\label{traverse_lock_b}	
	\end{subfigure}
	\caption{The \lstinline{traverse} method of the lock-coupling template algorithms}
	\label{traverse_lock}
\end{figure}

To utilize the template methodology for concurrent data structure implementations in C, we have implemented the the \lstinline{traverse} function in C, as depicted in Figure \ref{traverse_lock_a}. The \lstinline{pn} struct includes two fields, namely \lstinline{p} and \lstinline{n}, which respectively represent the current and next node. Furthermore, we require a target key \lstinline{k}, and a dummy node's value \lstinline{value} (referred to as 'dummy' because it is not actuallly used in \lstinline{traverse}). We include this dummy node to maintain consistency with other implementations. Our version of the \lstinline{traverse} function closely resembles the one written in an ML-like language since it relies on a helper function called \lstinline{findnext}. The purpose of \lstinline{findnext} is to determine the next node to be visited during the traversal. In our implementation, the \lstinline{traverse} function stops when it reaches the base case, which occurs when the current node is \lstinline{NULL}. Alternatively, it terminates if \lstinline{findnext} fails to find a next node that includes the key value $k$ in the set of key values. The \lstinline{traverse} function returns a Boolean value indicating whether it has successfully found the next node during traversal. If it fails to do so, it returns false, which other functions can use to make decisions. For instance, in a binary search tree, a node with a value of 20 would have a left child with a set of key values consisting of values less than 20, and a right child with one greater than 20. If we want to insert a new node with the target key $k$ into the tree, we can utilize the \lstinline{traverse} function to traverse the tree. When \lstinline{traverse} fails to find the next node (either to the left or right), it implies that it has reached a node with a target key $k$ that equals the node's value. At this point, we just have to modify the value of that node to \lstinline{value}. When \lstinline{findnext} arrives at a new node, it acquires a lock for that node, as is the nature of hand-over-hand locking. It then releases the lock for the current node, ensuring that the link between the two nodes is not removed or rearranged while traversing it.

\subsection{Give-Up}
The give-up template is another algorithm for implementing concurrent search in a tree data structure. Unlike the lock-coupling template, which maintains locks during traversal between nodes, the give-up template only acquires locks when loading or storing to a node. This reduces the amount of time that locks are held and makes the algorithm more scalable in highly-contended scenarios. Each node has the range field that serves as a reference point for threads to determine if they are still on the correct path. If a thread finds that it has deviated from the expected path, it abandons the node by releasing the lock and returns to the root node to start again. This behavior aligns with the give-up template algorithm's approach, wherein threads terminate their search upon encountering obstacles or deviating from the expected path.

@@

Figure \ref{traverse_giveup_b} depicts the give-up template algorithm, which is presented in an ML-like language and described in \cite{shasha_1988}. 
The give-up template algorithm uses a helper function called \lstinline{inRange} in addition to the \lstinline{findNext} function discussed earlier in the lock-coupling section. The purpose of the \lstinline{inRange} function is to determine if the key value $k$ is within the range of node $n$. If the key value is outside the node's range, the search is terminated, and the thread is restarted.
